NanoMaLy

### 2015-11-18 paper/pldi16-submission/semantics.tex
minor tweaks to semantics

\paragraph{Narrowing Types} The procedure $\force{e}{t}{\vsu}$ takes as
input an term \texttt{e} and refines it to have the type \texttt{t}
by yielding a pair of either the same expression and a value substitution
the refines the holes in \texttt{e} to ensure it has type \texttt{t},
or yields the stuck term if no such refinement is possible:
%
$$
\begin{array}{lcl}
% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
% \forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
\force{\ehole{i}}{t}{\vsu} & \defeq & \hspace{-1ex} \begin{array}{l}
    \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \pair{\vsu(i)}{\emptysu}\ \mathtt{else} \\
         \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}}
  \end{array} \\
\force{v}{\ehole{}}{\vsu}  & \defeq & \pair{v}{\emptysu} \\
\force{n}{\tint}{\vsu}     & \defeq & \pair{n}{\emptysu} \\
\force{v}{\tint}{\vsu}     & \defeq & \pair{\stuck}{\emptysu} \\
\force{b}{\tbool}{\vsu}    & \defeq & \pair{b}{\emptysu} \\
\force{v}{\tbool}{\vsu}    & \defeq & \pair{\stuck}{\emptysu} \\
\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\efun{x}{e}}{\emptysu} \\
\force{v}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\stuck}{\emptysu} \\
\end{array}
$$

@@

\paragraph{Primitive Reductions}
Primitive reduction steps -- addition, if-elimination, and
function application -- uses \forcesym to ensure that values have the
appropriate type (and that holes are instantiated) before continuing the
computation. Importantly, beta-reduction \emph{does not} type-check its
argument, it only ensures that the value being applied is a function.

@@

\paragraph{Witness Generality} We can now define
We now show that our evaluation relation instantiates holes in a
\emph{general} manner. That is, given a function $f$, if we have
$\steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su}$, then for every
possible input type $t$, we can find a value $\hastype{v}{t}$
such that $\eapp{f}{v}$ goes wrong.

@@

\paragraph{Dynamic Type Inference}
We can think of the evaluation of \eapp{f}{\ehole{}} as (dynamically)
inferring a \emph{partial type} --- a type that may contain holes ---
for $f$.
%
We can extract this type from evaluation traces.
%
Formally, we say that if \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\su},
then the \emph{partial type of} $f$ upto $\trace$, written \ptype{\trace}{f},
is \tfun{\typeof{\subst{\su}{\ehole{}}}}{\typeof{e}}.
%
%We will omit the subscript when we wish to refer to the final partial
%type, \ie\ at the step where the expression has been reduced to a value
%(or stuck.)

@@

\paragraph{Preservation}
Given these two definitions, we show that each evaluation step
refines the partial type of $f$, \ie preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \pair{\eapp{f}{\ehole{}}}{\emptysu},\ldots$ and
$\trace' \defeq \trace, \pair{e'}{\vsu'}$ (\ie $\trace'$ is single
step extension of $\trace$)
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
\end{lem}
\begin{proof}
  By induction on the length of $\trace$ and case analysis on the evaluation rules.
  %
  Note that all rules preserve partial types with the exception of when
  \forcesym is called on a hole, which case we instantiate the hole with
  a concrete value.
  %
  But \hastype{\ehole{}}{\thole{}}, which is compatible with any type.
\end{proof}

\paragraph{Narrowing}
%
Furthermore, only a call to \forcesym can change the partial type of $f$.
%
\begin{lem}
\label{lem:force-inst}
If
$\trace \defeq \pair{\eapp{f}{\ehole{}}}{\emptysu}, \ldots,\pair{e}{\vsu}$
and
$\trace' = \trace, \steps{e}{\vsu}{e'}{\vsu'}$
and
$\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
then the final step $\steps{e}{\vsu}{e'}{\vsu'}$ invokes \forcesym.
\end{lem}

@@

\paragraph{Incompatible Values Are Wrong}
%
We show that \emph{any} value that is \emph{incompatible} with
the partial type upto $\trace$ will cause $f$ to get stuck
in \emph{at most} $k+1$ steps where $k$ is the length of the
trace $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  If \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\vsu},
     not \tcompat{t}{\ptype{\trace}{f}} and
     \hastype{v}{t}
  then
     \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
\end{lem}

\begin{proof}
  By induction on $k$, the length of $\trace$.
  Let $\tfun{s_k}{\_} = \ptype{\trace}{f}$.
  Suppose \hastype{v}{\tincompat{t}{s_{\trace'}}}, we
  will show that \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su'}
  in at most $k+1$ steps:
  \begin{description}
  \item[Case \tincompat{t}{s_k}:]
    The inductive hypothesis applies.
  \item[Case \tcompat{t}{s_k} but \tincompat{t}{s_{k+1}}:]
    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
    $k+1$.
    %
    A case analysis of the applicable rules shows that \forcesym cannot
    have succeeded.
  \end{description}
\end{proof}

@@

\paragraph{Generating Witnesses}
% Procedure \texttt{genWitness}}
%
Finally, Figure~\ref{fig:algo-gen-witness} summarizes the overall
implementation of our search for witnesses, in procedure @genWitness@
which takes as input a bound @n@ and a list of top level binders, and
returns a list of witness expressions $e$ which demonstrate how the input
program gets stuck.
%
The search proceeds as follows.
%
@(1)@ we invoke @genArgs@ to determine the number of parameters required
by the (curried) target function.
%
@(2)@ we obtain a closed expression $e$ by closing the target with the
computed holes.
%
@(3)@ we take the first @n@ traces returned by @eval@ on the closed $e$
and
@(4)@ we extract the substitutions corresponding to the $\stuck$ traces,
and use them to return the list of witnesses.
%
We obtain the following as a corrollary of Theorem~\ref{thm:soundness}:

>>>>

\paragraph{Narrowing Types} The procedure $\force{e}{t}{\vsu}$ takes as
input an term \texttt{e} and refines it to have the type \texttt{t} by
yielding a pair of either the same expression and a value substitution
the refines the holes in \texttt{e} to ensure it has type \texttt{t}, or
yields the stuck term if no such refinement is possible. In the case
where \texttt{e} is a hole, it first checks in the given $\vsu$ to see
if the hole has already been instantiated and, if so, returns the
existing instantiation:
%
$$
\begin{array}{lcl}
% \multicolumn{3}{l}{\forcesym \ :: \  (e, t) \rightarrow \pair{e}{\vsu}} \\
% \forcesym            & ::     & (e, t) \rightarrow \pair{e}{\vsu} \\
\force{\ehole{i}}{t}{\vsu} & \defeq & \hspace{-1ex} \begin{array}{l}
    \mathtt{if}\ i \in \vsu\ \mathtt{then}\ \pair{\vsu(i)}{\emptysu}\ \mathtt{else} \\
         \elet{v}{\gen{t}}{\pair{v}{\ehole{i} \mapsto v}}
  \end{array} \\
\force{v}{\ehole{}}{\vsu}  & \defeq & \pair{v}{\emptysu} \\
\force{n}{\tint}{\vsu}     & \defeq & \pair{n}{\emptysu} \\
\force{v}{\tint}{\vsu}     & \defeq & \pair{\stuck}{\emptysu} \\
\force{b}{\tbool}{\vsu}    & \defeq & \pair{b}{\emptysu} \\
\force{v}{\tbool}{\vsu}    & \defeq & \pair{\stuck}{\emptysu} \\
\force{\efun{x}{e}}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\efun{x}{e}}{\emptysu} \\
\force{v}{\tfun{\thole{}}{\thole{}}}{\vsu} & \defeq & \pair{\stuck}{\emptysu} \\
\end{array}
$$

@@

\paragraph{Primitive Reductions}
Primitive reduction steps -- addition, if-elimination, and
function application -- use \forcesym to ensure that values have the
appropriate type (and that holes are instantiated) before continuing the
computation. Importantly, beta-reduction \emph{does not} type-check its
argument, it only ensures that the value being applied is a function.

@@

\paragraph{Witness Generality}
We now show that our evaluation relation instantiates holes in a
\emph{general} manner. That is, given a function $f$, if we have
$\steps{\eapp{f}{\ehole{}}}{\emptysu}{\stuck}{\su}$, then for every
possible input type $t$, we can find a value $\hastype{v}{t}$
such that $\eapp{f}{v}$ goes wrong.

@@

\paragraph{Dynamic Type Inference}
We can think of the evaluation of \eapp{f}{\ehole{}} as (dynamically)
inferring a \emph{partial type} --- a type that may contain holes ---
for $f$.
%
We can extract this type from evaluation traces.
%
Formally, we say that if \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\su},
then the \emph{partial type} of $f$ upto $\trace$, written \ptype{\trace}{f},
is \tfun{\typeof{\subst{\su}{\ehole{}}}}{\typeof{e}}.
%
%We will omit the subscript when we wish to refer to the final partial
%type, \ie\ at the step where the expression has been reduced to a value
%(or stuck.)

@@

\paragraph{Preservation}
Given these two definitions, we show that each evaluation step
refines the partial type of $f$, \ie preserves type compatibility.
%
\begin{lem}
\label{lem:refine-partial}
If $\trace \defeq \pair{\eapp{f}{\ehole{}}}{\emptysu},\ldots$ and
$\trace' \defeq \trace, \pair{e'}{\vsu'}$ (\ie $\trace'$ is single
step extension of $\trace$)
%
%The partial type of $f$ upto $\trace$ is compatible
%with the partial type upto $\trace'$, \ie\
%
then \tcompat{\ptype{\trace}{f}}{\ptype{\trace'}{f}}.
\end{lem}
\begin{proof}
  By induction on the length of $\trace$ and case analysis on the evaluation rules.
  %
  Note that all rules preserve partial types with the exception of when
  \forcesym is called on a hole, in which case we instantiate the hole with
  a concrete value.
  %
  But \hastype{\ehole{}}{\thole{}}, which is compatible with any type.
\end{proof}

\paragraph{Narrowing}
%
Furthermore, only a call to \forcesym can change the partial type of $f$.
%
\begin{lem}
\label{lem:force-inst}
If
$\trace \defeq \pair{\eapp{f}{\ehole{}}}{\emptysu}, \ldots,\pair{e}{\vsu}$
and
$\trace' \defeq \trace, \step{e}{\vsu}{e'}{\vsu'}$
and
$\ptype{\trace}{f} \neq \ptype{\trace'}{f}$,
then the final step $\step{e}{\vsu}{e'}{\vsu'}$ invokes \forcesym.
\end{lem}

@@

\paragraph{Incompatible Values Are Wrong}
%
We show that \emph{any} value that is \emph{incompatible} with
the partial type upto $\trace$ will cause $f$ to get stuck
in \emph{at most} $k+1$ steps where $k$ is the length of the
trace $\trace$.
%
\begin{lem}
\label{lem:k-stuck}
  If \steptr{\trace}{\eapp{f}{\ehole{}}}{\emptysu}{e}{\vsu},
     \tincompat{t}{\ptype{\trace}{f}}, and
     \hastype{v}{t},
  then
     \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su}.
\end{lem}
\ES{self-todo: take a close look at the $\trace$'s vs k's}
\begin{proof}
  By induction on $k$, the length of $\trace$.

  Let $\tfun{s_{\trace}}{\_} = \ptype{\trace}{f}$.
  Suppose \hastype{v}{\tincompat{t}{s_{\trace'}}}, we
  will show that \steps{\eapp{f}{v}}{\emptysu}{\stuck}{\su'}
  in at most $k+1$ steps:
  \begin{description}
  \item[Case \tincompat{t}{s_{\trace}}:]
    The inductive hypothesis applies.
  \item[Case \tcompat{t}{s_{\trace}} but \tincompat{t}{s_{\trace'}}:]
    By Lemma~\ref{lem:force-inst} we must have called \forcesym at step
    $k+1$.
    %
    A case analysis of the applicable rules shows that \forcesym cannot
    have succeeded.
  \end{description}
\end{proof}

@@

\paragraph{Generating Witnesses}
% Procedure \texttt{genWitness}}
%
Finally, Figure~\ref{fig:algo-gen-witness} summarizes the overall
implementation of our search for witnesses, in the procedure @genWitness@,
which takes as input a bound @n@ and a list of top level binders, and
returns a list of witness expressions $e$ that demonstrate how the input
program gets stuck.
%
The search proceeds as follows.
%
@(1)@ we invoke @genArgs@ to determine the number of parameters required
by the (curried) target function.
%
@(2)@ we obtain a closed expression $e$ by closing the target with the
computed holes.
%
@(3)@ we take the first @n@ traces returned by @eval@ on the closed $e$
and
@(4)@ we extract the substitutions corresponding to the $\stuck$ traces,
and use them to return the list of witnesses.
%
We obtain the following as a corrollary of Theorem~\ref{thm:soundness}:
